---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gepabds

<!-- badges: start -->
<!-- badges: end -->

The goal of gepabds is to provide tools for analyzing and visualizing gene expression patterns in single-cell RNA-seq data. It enables per-group expression summaries, gene specificity scoring, and visualization of marker genes using datasets.

## Installation

You can install the development version of gepabds from [GitHub](https://github.com/) with:

``` r
install.packages("remotes")
remotes::install_github("lvmedha/gepabds")
```

```{r example}
library(gepabds)
library(SingleCellExperiment)

# load example dataset
data(example_se)

# genes to explore
genes <- c("INS", "GCG", "SST", "PRSS1")

# 1. Expression statistics
stats <- compute_expr_stats(example_se, genes)
head(stats)

# 2. Gene specificity
spec <- compute_gene_specificity(example_se, genes)
spec

# 3. Long-format data for plotting
expr_long <- build_expr_long(example_se, genes)

# 4. Visualizations
plot_violin(expr_long)

plot_scatter(example_se, "INS", "GCG")
```

## Example Data

This package includes a subset of the Muraro human pancreas single-cell RNA-seq dataset (`example_se`) for testing purposes.
