---
title: "Getting Started with gepabds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with gepabds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(gepabds)
library(SummarizedExperiment)
```

Introduction 

The gepabds package provides tools for analyzing gene expression data stored in SummarizedExperiment objects.

This vignette shows how to compute expression statistics for selected genes.

Loading the Data:

```{r}

# load data
data(example_se)

example_se
```

Exploring the data:

```{r}
# Explore data
# View gene names
rownames(example_se)

# View sample metadata
colData(example_se)
```

Genes of interest

This step selects a small subset of genes from the dataset to demonstrate downstream analysis.

 In real analyses, genes of interest are typically selected based on biological relevance or statistical criteria, such as:

high mean expression across samples
high variability between conditions
known involvement in a pathway or disease process

```{r}
# genes of interest
genes_to_use <- rownames(example_se)[1:5]
genes_to_use
```


Computing Expression Stats

This function calculates summary statistics for each gene, including:
- mean expression across samples
- variance (how variable expression is)
- optionally filtered results for selected genes

These metrics help prioritize genes that are biologically informative.

```{r}

# compute expression statistics

result <- compute_expr_stats(example_se, genes = genes_to_use)

result
```

Interpreting the results:

Each row corresponds to a gene. Genes with higher mean expression are more 
abundant across samples, while higher variance suggests condition-specific regulation.

In downstream analyses, we often focus on genes that are both highly expressed 
and variable, as these are more likely to be biologically meaningful.
